---
title: "R for Data Science"
subtitle: 'Chapter 5 Exercises'
author: "Robert Kirk DeLisle"
output: html_document
---

```{r include=FALSE}
library(tidyverse)
library(nycflights13)
```

## Chapter 5 - Data Transformations

### 5.2.4 Exercises

**1. Find all flights that: **

    **1.  Had an arrival delay of two or more hours**
    **2.  Flies to Houston (`IAH` of `HOU`)**
    **3.  Were operated by United, American, or Delta**
    **4.  Departed in summer (July, August, and September)**
    **5.  Arrived more than two hours late, but didn't leave late**
    **6.  Were delayed by at least an hour but mad up over 30 minutes in flight**
    **7.  Departed between midnight and 6 am (inclusive)** 
    
```{r}
# 1
f <- filter(flights, arr_delay >= 2)
range(f$arr_time)

# 2
f <- filter(flights, dest %in% c('IAH','HOU'))
unique(f$dest)

# 3
f <- filter(flights, carrier %in% c('AA','UA','DL'))
unique(f$carrier)

# 4
f <- filter(flights, month %in% c(7,8,9))
unique(f$month)

# 5
f <- filter(flights, arr_delay > 120 & dep_delay <= 0)
head(f)
  
# 6
f <- filter(flights, dep_delay >= 60, dep_delay - arr_delay >= 30)
head(f)

# 7
f <- filter(flights, dep_time == 2400 | dep_time <= 600)
range(f$dep_time)
```

**2. Another useful dplyr filter helper is `between()`.   What does it do?  Can you use it to simply the code needed to answer the previous challenges?**

From the help file:  This is a shortcut for x >= left & x <= right.

It can be used to change a few of the challenges from 1, above.  

```{r}
# 4
f <- filter(flights, between(month, 7, 9))
unique(f$month)

# 7 
f <- filter(flights, dep_time == 2400 | between(dep_time, 1, 600))
range(f$dep_time)
```

**3. How many flights have missing `dep_time`?  What other variables are missing?  What might these rows represent?**

Looking at the table, `dep_delay`, `arr_time`, `arr_delay`, `air_time` are all missing.  These are probably cancelled flights.

```{r}
nrow(f <- filter(flights, is.na(dep_time)))
```
 
**4. Why is `NA ^ 0` not missing?  Why is `NA | TRUE` not missing?  Why is `FALSE & NA` not missing?  Can you figure out a general rule?  (`NA * 0` is a tricky counterexample!)**

`NA ^ 0` - any number to 0 power is 1

`NA | TRUE` - the or operator (`|`) allows one `TRUE` value to override the `NA`

`FALSE & NA` - the & operator ('&') finds one `FALSE` and thus returns `FALSE`

Generally, when we consider the context of any operation, if the mathematical or logical result for any possible value of `NA` is computable, that value is returned.  `NA * 0` doesn't work becuase of potential imprecision in the value of the `NA`.
 
 ### 5.3.1 Exercises
 
 **1. How could you use `arrange()` to sort all missing values to the start?  (Hint: use `is.na()` ).**
 
```{r}
arrange(flights, desc(is.na(dep_time)))
```
 
 **2.  Sort `flights` to find the most delayed flights.  Find the flights that left the earliest.**
 
```{r}
arrange(flights, desc(dep_delay))
arrange(flights, dep_delay)
```

**3.  Sort `flights` to find the fastest (highest speed) flights.**

```{r}
arrange(flights, distance/air_time)
```

**4. Which flights travelled the farthest?  Which travelled the shortest?**

```{r}
arrange(flights, desc(distance))
arrange(flights, distance)
```

### 5.4.1 Exercises

**1. Brainstorm as many ways as possible to select `dep_time`, `dep_delay`, `arr_time`, and `arr_delay` from `flights`.**

```{r}
select(flights, starts_with('dep'), starts_with('arr'))
```

**2.  What happens if you include the name of a variable multiple times in a `select()` call?**

The repeated variable is only selected once.

```{r}
select(flights, dep_time, dep_time) 
```

**3.  What does the `one_of()` function do?  Why might it be helpful in conjunction with this vector?**

`one_of` is retired, but it appears to select all of the variables listed in the `vars` vector.  Help file suggests using `any_of` or `all_of` instead.

```{r}
vars <- c('year', 'month', 'day', 'dep_delay', 'arr_delay')
select(flights, one_of(vars))
```

**4.  Does the result of running the following code surprise you?  How do the select helpers deal with case by default?  How can you change that default?**

Apparently, the default is to ignore case.  The `contains()` function has a parameter called `ignore.case` which is `TRUE` by default.

```{r}
select(flights, contains('TIME'))
```

### 5.5.2 Exercises

**1.  Currently `dep_time` and `sched_dep_time` are convenient to look at, but hard to compute with because they're not really continuous numbers.  Convert them to a more convenient representation of number of minutes since mindnight.**

```{r}
f <- mutate(flights, deptime_min = as.integer(dep_time/100)*60 + 
                                   100*(dep_time/100 - as.integer(dep_time/100)),
                     arrtime_min = as.integer(arr_time/100)*60 +
                                   100*(arr_time/100 - as.integer(arr_time/100)))
head(select(f, one_of(c('dep_time', 'deptime_min', 'arr_time', 'arrtime_min'))))
```

**2.  Compare `air_time` with `arr_time - dep_time`.  What do you expect to see?  What do you see?  What do you need to do to fix it?**

```{r}
f <- select(flights, air_time, arr_time, dep_time)
head(mutate(f, delta = arr_time - dep_time))
```

The subtraction doesn't take into account that `air_time`, `arr_time` and `dep_time` are in `HHMM` format.  They need to be converted as done in 1, above.

**3. Compare `dep_time`, `sched_dep_time`, and `dep_delay`.  How would you expect those three number to be related?**

`dep_delay` should be the difference in minutes between `dep_time` and `sched_dep_time`.

```{r}
f <- select(flights, dep_delay, dep_time, sched_dep_time)
head(f)
```


**4.  Find the 10 most delayed flights using a ranking function.  How do you want to handle ties?  Carefully read teh documentation of `min_rank()`.**

By default, `min_rank()` assigned the minimum value of the tied values the same rank.  There is no way to adjust this.

```{r}
arrange(flights, desc(min_rank(dep_delay)))
```

**5.  What does `1:3 + 1:10` return?  Why?**

An error because the vectors are not of the same length.

**6.  What trigonometric function does R provide?**

sin, cos, tan, sinh, cosh, tanh, asin, acos, atan, asinh, acosh, atanh

### 5.6.7 Exercises

### 5.7.1 Exercises

