---
title: "R for Data Science"
subtitle: 'Chapter 4 Exercises'
author: "Robert Kirk DeLisle"
output: html_document
---

```{r include=FALSE}
library(tidyverse)
```

## Chapter 4 - Workflow: scripts

### 4.4 Exerices

**1. Why does this code not work?**

```{r}
my_variable <- 10
#my_varıable
```

The character used for 'i' in the second line is not consistent with that used in the first line.  Unicode character?  The two variables are essentially diferent names.

**2.  Tweak each of the following R commands so that they run correctly:**

```{r}
library(tidyverse)

# original code
# ggplot(dota = mpg) + 
#   geom_point(mapping = aes(x = displ, y = hwy))
# fliter(mpg, cyl = 8)
# filter(diamond, carat > 3)

ggplot(data = mpg) + # data was spelled with an 'o'
  geom_point(mapping = aes(x = displ, y = hwy))

filter(mpg, cyl == 8) # filter was misspelled, need '==' instead of '='
filter(diamonds, carat > 3) # diamonds lacked the 's'
```

**3. Press Alt + Shift + K.  What happens?  How can you get to the same place using the menus?**

On the Mac, Option + Shift + K give a table of keyboard shortcuts.  From the menu, Tools -> Keyboard Shortcuts Help.

