---
title: "R for Data Science"
subtitle: 'Chapter 11 Exercises'
author: "Robert Kirk DeLisle"
output: html_document
---

```{r include=FALSE}
library(tidyverse)
library(readr)
```

## Chapter 11 - Data import

### 11.2.2 Exercises

**1.  What function would you use to reaad a file where fields were separated with a "|"?**

`read_delim(file = <filename>, delim = '|')`

**2.  Apart from `file`, `skip`, and `comment`, what other arguments do `read_csv()` and `read_tsv()` have in common?**

`col_names`, `col_types`, `locale`, `na`, `quoted_na`, `quote`, `trim_ws`, `n_max`, `guess_max`, `progress`, `skip_empty_rows`

**3.  What are the most important arguments to `read_fwf()`?**

This answer depends on the file being read, but `col_positions` is probably extremely important as it determines where columns start.

**4.  Sometimes strings in a CSV file contain commas.  To prevent them from causing problems they need to be surrounded by a quoting character, like `"` or `'`.  By default `read_csv()` assumes that the quoting character will be `"`.  What argument to `read_csv()` do you need to specify to read the following text into a data frame?**

`"x,y\n1,'a,b'"`

```{r}
tib <- readr::read_csv("x,y\n1,'a,b'", quote="'")
tib
```

**5.  Identify what is wrong with each of the following inline CSV files.  What happends when you run the code?**

```{r}
read_csv("a,b\n1,2,3\n4,5,6") 
# two columns in header, 3 columns in data
# data extracted by row

read_csv("a,b,c\n1,2\n1,2,3,4")
# data extracted by row, so 3 columns in header
# 2 and 4 columns in each data row

read_csv("a,b\n\"1")
# only one data element in data row

read_csv("a,b\n1,2\na,b")
# data is converted to <chr> since second data row and characters

read_csv("a;b\n1;3")
# delimiter is ; - use read_delim() and set delim = ';'
```

### 11.3.5 Exercises

**1.  What are the most important argumetns to `locale()`?**

The various arguments all specification of the format for dates, times, grouping marks, time zones, and encodings based upon specific needs.  All of them will be important to some extent depending on the nature of the input data.

**2.  What happens if you try and set `decimal_mark` and `grouping_mark` to the same character?  What happens to the default value of `grouping_mark` when you set `decimal_mark` to ","?  What happens to the default value of `decimal_mark` when you set the `grouping_mark` to "."?**

```{r}
# an error occurs when deimal_mark and grouping_mark are the same - they must be different
parse_number("123,456,789", 
             locale = locale(decimal_mark = ',', grouping_mark = ','))

# presumably, grouping mark is set to '.' but I cannot find specific documentation of that
parse_number("123.456,789", 
             locale = locale(decimal_mark = ','))

# Again, presumably it is set to ',' but I cannot find specfic documentation
parse_number("123.456,789", 
             locale = locale(grouping_mark = '.'))
```

**3.  I didn't discuss the `date_format` and `time_format` options to `locale()`.  What do they do?  Contruct an example that shows when they might be useful?**

These options will allow specification of time and data formats in input data.

```{r}
parse_datetime("2010/01/01 12:00 US/Central", "%Y/%m/%d %H:%M %Z")
```

**4.  If you live outside the US, create a new locale object that encapsulates the setting for the types of file you read most commonly.**

```{r}
# Example with , as decimal mark and dates written as YY.MM.DD.

parse_date('20.09.15', locale = locale(date_format = '%y.%m.%d', decimal_mark = ',', tz = 'GMT'))
```

**5.  What's the difference between `read_csv()` and `read_csv2`?**

`read_csv()` assumes comma delimiters, and `read_csv2()` assumes semicolon delimiters.

**6.  What are teh most common encodings used in Europe?  What are the most common encodings used in Asia?  Do some googling to find out.**

Western Europe - ISO 8859-1
Western and Central Europe - ISO 8859-2
Greek - ISO 8859-7
etc.

Traditional Chinese - Big5
Simplified Chinese - GB18030
Japanese - Shift-JIS, EUC-JP

**7.  Generate the correct format string to parse each of the following dates and times.**

```{r}
d1 <- "January 1, 2010"
parse_date(d1, locale = locale(date_format = '%B %d, %Y'))

d2 <- "2015-Mar-07"
parse_date(d2, locale = locale(date_format = '%Y-%b-%d'))

d3 <- "06-Jun-2017"
parse_date(d3, locale = locale(date_format = '%d-%b-%Y'))

d4 <- c("August 19 (2015)", "July 1 (2015)")
parse_date(d4, locale = locale(date_format = '%B %d (%Y)'))

d5 <- "12/30/14" # Dec 30, 2014
parse_date(d5, locale = locale(date_format = '%m/%d/%y'))

t1 <- "1705"
parse_time(t1, locale = locale(time_format = '%H%M'))

t2 <- "11:15:10.12 PM"
parse_time(t2, locale = locale(time_format = '%I:%M:%OS %p'))
```

