---
title: "R for Data Science"
subtitle: 'Chapter 10 Exercises'
author: "Robert Kirk DeLisle"
output: html_document
---

```{r include=FALSE}
library(tidyverse)
library(nycflights13)
library(ggstance)
library(lvplot)
```

## Chapter 10 - Tibbles

### 10.5 Exercises

**1.  How can you tell if an object is a tibble?**

Printing a data frame produces simply a table of data.  Printing a tibble starts with the line *# A tibble: <number of rows> x <number of columns>*.  Additionallly, th tibble lists the column data type, row numbers in light text at the beginning of each row, and a closing line telling how many additional rows are present in the tibble.

**2.  Compare and contrast the following operations on a `data.frame` and equivalent tibble.  What is different?  Why might the default data frame behaviours cause you frustration?**

```{r}
df <- data.frame(abc = 1, xyz = "a")
# no major differences in creation of data frame or tibble

df$x
# the data frame will match the `xyz` column but the tibble will not
# this could lead to unexpected behavior if the use of 'x' was accidental - an error should occur, as with a tibble

df[, "xyz"]
# The data frame will return a vector of one element
# The tibble will return a 1x1 tibble

df[, c("abc", "xyz")]
# both will return a new data frame or tibble with the specified columns
```

**3.  If you have the name of a variable stored in an object, w.g. `var <- "mpg"`, how cna you extract the reference variable from a tibble?**

```{r}
tb <- tibble(mtcars)
var <- "mpg"
tb[var]
dplyr::select(tb, var)
```

**4. Practice referring to non-syntactic names in the following data frame by:

1.  Extracting the variable called `1`.
2.  Plotting a scatterplot of `1` vs `2`.
3.  Creating a new column call `3` which is `2` divided by `1`.
4.  Renaming the columns to `one`, `two`, and `three`.**

```{r}
annoying <- tibble(
  `1` = 1:10,
  `2` = `1` * 2 + rnorm(length(`1`))
)

annoying$`1`

plot(annoying$`1`, annoying$`2`)

annoying$`3` = annoying$`2`/annoying$`1`

annoying <- dplyr::rename(annoying, 'one' = `1`, 
                                    'two' = `2`,
                                    'three' = `3`)
annoying
```

**5. What does `tibble::enframe()` do?  When might you use it?**

`enframe()` converts named atomic vectors or lists to one- or two-column data frames. For a list, the result will be a nested tibble with a column of type list. For unnamed vectors, the natural sequence is used as name column.

`enframe()` would be useful to create a type of assiciative array.  The value in the `name` column operates as the key to the data object in the `volue` column.

**6.  What option controls how many additional column names are printed at the footer of a tibble?**

The `width` option of `tibble.print()` or `options(tibble.width = <#>)` with `<#>` being the number of additional columns. `options(tibble.width = Inf)` will print all columns.



